Direct observation of Feshbach enhanced \swave scattering of fermions

### 2015-05-05 draft1.tex
made the tex file run on home laptop
Went through almost all of second round of Ian's comments. Haven't fixed the figures or some notes starred on the paper.

\begin{document}
\newcommand{\swave}[0]{$\it{s}$-wave }
\newcommand{\pwave}[0]{$\it{p}$-wave }
\newcommand{\K}{ $^{40}\rm{K}$}
\newcommand{\Rb}{ $^{87}\rm{Rb}$}
\newcommand{\us}{$\mu \rm{s}$}

@@

\begin{abstract}
We directly measured the \swave{} scattering cross-section of \K atoms accross a Feshbach resonance. We collided a pair of degenerate clouds and imaged the scattered atoms. Owing to their low density, few atoms scattered, even near the resonance. To optimize signal to noise, we develop techniques to interpret absorption imaging in the regime where the optical intensity changes dramatically as light traverses the cloud, and recoil induced 
detuning corrections are significant. We applied these techniques to our \swave{} scattering data and extracted the resonant magnetic
field value. These imaging techniques are generally applicable, and can be used to observe effective \pwave{} scattering in the presence of spin-orbit coupling in a spin polarized  Fermi gas.

@@

\section{Introduction}
Feshbach resonances are a stand-by technique for tuning the interaction strength in ultracold atomic gases. Even weak interactions play a crucial role in the physics of atomic Bose-Einstein Condensates (BECs), giving rise to, for example, their characteristic Thomas-Fermi density profiles. The effects of interactions in Fermi gases, however, are harder to observe. The density of Fermi clouds is typically reduced by a factor of $~10^3$ from that of BECs, making it necessary to enhance the strength of interactions in order to observe them. Feshbach resonances alter the inter-atomic scattering length via the magnetic field. Not only do they enable detection of interactions, but they make it possible to tune interactions from attractive to repulsive, allowing for the phase transition from the BCS to BEC regime at sufficiently low temperatures \cite{RegalThesis}. 
\par A Feshbach resonance occurs when a bound molecular state of few atoms energetically approaches the free atomic state \cite{Chin10}. The energy of the atomic states is defined by their hyperfine state and tunable with a bias magnetic field. The Feshbach resonance can thus be approached by changing the bias field. The effect of the resonance on the scattering length between two free atoms in the appropriate hyperfine states is
\begin{equation}
a(B)=a_{bg}\left(1-\frac{\Delta}{B-B_0}\right),
\label{feshbachEq}
\end{equation}
where $a$ is the scattering length, $a_{bg}$ is the scattering length away from the resonance, $\Delta$ is the width of the resonance, and $B_0$ is the field value at which the resonance occurs. Note that the scattering length tends to infinity from either side of the resonance.
\par To use Feshbach resonances as a tool it  is necessary to have a good measurement of the parameters in the above equation.  The exact value of the resonant field value $B_0$ is impossible to calculate analytically and must be estimated via numerical models \cite{Lysebo09, Gao11} or determined experimentally. Some experimental techniques that have been used to characterize Feshbach resonances include: the observation of loss due to three-body inelastic scattering, re-thermalization timescales, or anisotropic expansion, all of which infer the elastic scattering cross section from collective behavior of the cloud \cite{Regal03,OHara02,Monroe93}. 
\par Here we present a new technique for measuring the location and width of a Feshbach resonance. We collide a pair of ultra-cold Fermi gases and directly image the resulting \swave scattering halo as a function magnetic field strength. This allows us to observe the enhancement in scattering without relying on proxy effects. We measured the fraction of atoms scattered during the collision, and from this fraction deduced the resonant magnetic field  and width of the resonance.
%\par The techniques developed in this experiment for observing Fermion scattering can be extended to engineering higher order partial wave interactions, as has been done for bosons \cite{Williams2012}.
\par Our Fermi gases are so dilute that even with the resonant enhancement of the scattering cross section, only a small fraction of the atoms scatter. This makes direct detection of \swave scattering halos difficult due to detection uncertainty, which disproportionately affects regions of low atomic density. To optimize the signal to noise for low atom numbers, we utilized a non standard imaging regime. In this regime, simulation was necessary for an accurate interpretation of the images. We performed these simulations and used the results to extract the atom number and the scattered fraction from our images.
\par This paper is in two parts. In the first, we study absorption imaging in the presence of significant time-dependent Doppler shift and show how we use our results to interpret data. In the second, we describe our\swave scattering experiment and extract a measure of the location of the Feshbach resonance in \K.

@@

Every experiment requires a detection scheme. In ultracold atomic experiments, detection relies on imaging laser light that has interacted in some way with the atomic cloud. The most common such imaging method, and the one we use in our experiments, is absorption imaging. Absorption imaging utilizes some excited to ground state transition between two levels of the atom. Such an atomic transition has an energy difference $E_0$, an associated frequency $\omega_0 = E_0/\hbar$, and a natural transition linewidth $\Gamma$. When interacting with a laser light field an atom will scatter photons: if will absorb a photon from the light field and move up to its excited state, and then re-emit the photon in a random direction and decay back to its ground state. The  rate at which this scattering occurs, given a laser intensity $I$, is given by

@@

where $I_{sat}$ is the saturation intensity, the intensity at which the atom is equally likely to be in the ground and excited state, and $\delta$ is the detuning, the difference between the resonant transition frequency $\omega_0$ and the frequency of the laser light $\omega_L$.  
\par To obtain an absorption image, one shines an on or near resonant probe beam ($\delta\approx0$) onto the atomic cloud. Some of the light is scattered by the atoms, and the part of the beam that made it through the cloud is imaged onto a camera, as seen in Fig. \ref{fig:absorptionIntor}a (top). Then, the atoms are allowed to leave the trap, and the probe light is shined directly at the camera to calibrate the intensity of light that the atoms saw (bottom). The intensity in the second camera image is called the initial intensity, $I_{0}$, as it is assumed to be the same as the intensity in the first beam before interaction with atoms. The intensity in the first image is called the final intensity, $I_f$.  

@@

\caption{Absorption imaging. a. First, a near resonant probe light is shined on the atoms, and the shadow of that light is imaged on the camera. Then, the atoms are allowed to escape their trap and an image of just the probe light is taken. b.  As the probe beam, travels through the atomic cloud, some of its power is absorbed, and the intensity seen by atoms further along the imaging direction $z$ is lowered.  c. An atomic cloud illuminated by a probe light field absorbs photons from the probe and re-emits them in arbitrary directions. This process results in a net velocity of the cloud in the direction of the probe light as well as diffusive spreading in the transverse directions.  }  

@@

\par The observed intensity in the two images can be used to infer the number of atoms that the light encountered. To do this, consider what happens to the light as it travels along the imaging axis $z$ through a column of atoms. In general, the atomic density is three dimensional, $\rho(x,y,z)$. For the purposes of this discussion, let us focus on a single pixel of the camera, and thus a single value of $I_0$ and $I_f$ and a single column of atoms, $\rho(z)$. We will not be able to infer the entire atomic distribution, but for each pixel we will be able to deduce the total 2-d atomic density, $n = \int \rho\left(z\right) \,\mathrm{d}z$. As the light travels through a column of atoms, each atom will scatter light according to Eq. (\ref{eq:scatrate}). Therefore, the atoms further along the imaging axis $z$ will see a reduced light intensity, as seen in Fig. \ref{fig:absorptionIntor}b. The amount of light lost to scattering as a function of $z$ is given by

@@

where $\sigma_0$ is the resonant scattering cross section. We have assumed $\delta=0$. 
 This equation can be easily integrated over $z$ to obtain  \cite{Reinaudi07}

@@

In the limit where the probe intensity is much smaller than the saturation intensity, $I_0\ll I_{sat}$, this becomes simply $-\ln \left(I_f/I_0\right)$. This quantity is called the optical depth, or OD. Since this is the simplest possible model that relates the observed intensities to the column density $n$, we call this quantity $OD_0$. Once the probe intensity becomes comparable to the saturation intensity, the second term in Eq. (\ref{eq2}) becomes significant. We can then call this the optical density corrected for high probe intensity, or

@@

\par When an atom absorbs a photon from the laser light field and gets excited to a higher energy level, by conservation of momentum it must also acquire a velocity in the direction of the light field. This is called the recoil velocity, and it is given by $v_r=\hbar k/m$, where $k$ is the wavenumber of the light field and $m$ is the atomic mass. When the photon is subsequently emitted and the atom returns to its ground state, it is emitted in an arbitrary direction. Thus, the momentum lost due to emitting a photon is on average zero. The variance, however, is not zero, allowing the atoms to acquire some momentum transverse to the laser field. We will ignore this correction, but the effect of this on the atomic cloud is picture in Fig. \ref{fig:absorptionIntor}c. 
\par Once the atoms absorb enough photons, they acquire a substantial velocity in the direction of the light field, and this velocity Doppler shift them from the light field itself. Thus, an atom that absorbs $N$ photons acquires a velocity of $N v_r$ and a detuning of $\delta=k N v_r$ from the probe beam. Therefore, even if the probe beam is initially on resonance with the atomic transition, we cannot neglect the detuning term in the scattering rate as time goes on. Furthermore, this detuning will vary both with imaging time $t$ and with distance along the propagation direction $z$ (Fig. \ref{fig:detunedBlobs}). Thus, the intensity lost to the atoms also acquires a time dependence: 

@@

\frac{dI(t,z)}{dz}=\sigma_0 \rho \frac{I(t,z)}{1+(2\Gamma/\delta(t,z))^2 +I(t,z)/I_{sat}}, \label{eq3}

@@

where the detuning $\delta$ is given by 

@@

\delta(t,z)=\frac{v_r}{\hbar c \rho}\int_0^t \frac{dI(z,\tau)}{dz}\,\mathrm{d}\tau. \label{eq4} 

@@

In this case, one can no longer obtain a straightforward relation between the atom number and the intensities in the two absorption images.

@@

	\includegraphics*{figure1.pdf}
\caption{Distribution of generalized detuning $\Delta=\frac{2\delta}{\Gamma}$ across an atomic cloud of $^{40}K$ for three different imaging times, as obtained by numerical simulation.}  

@@

\par One can consider this equation perturbatively in time and obtain corrections to second order in imaging time, $OD_2$  \cite{LJLthesis}. However, the perturbative treatment breaks down shortly after the zeroth order approximation of Eq. (\ref{eq2}) (Fig. \ref{fig:ODcorrections}). In order to adequately correct for the recoil induced detuning of the atoms, we must simulate the process and obtain numerical predictions for $I_f$ given a certain imaging time, atomic density, and probe intensity. 

@@

\par In the following, we describe two versions of this simulation. First, we take the simplistic approach that the on-axis distibution of atoms does not change dramatically durning the imaging time and can be treated statically. We test this approach in known limits and then check the validity of the static assumption. It turns out that, for realistic input parameters, this assumption is grossly incorrect. So, we take a slightly more sophisticated approach and allow the atoms to move within the cloud during the imaging time. This allows us to simulate the phase space evolution of atoms subjected to probe light. However,   
 we find that  while the atomic trajectories are wildly different, the predicted ODs only vary on the 0.5$\%$ level between the two models.

\subsection{Stationary atom model}
In order to solve eqs. [\ref{eq3}]-[\ref{eq4}], we start with an input 1-d distribution of atom densities $\rho(z)$, usually Gaussian in shape. We divide the cloud into spatial bins (the bin size was chosen such that decreasing the bin size further produced less than a 0.001$\%$ difference in the result).  In this approximation, we keep the number of atoms in each bin constant.  The algorithm used is shown in alg. [\ref{algorithm1}]. We call the optical depth obtained from this algorithm $OD_{corr1}$.

\begin{algorithm}
\caption{Stationary atom model}
\label{algorithm1}
\begin{algorithmic}
\STATE $I[n=0,t]=I_0$ \COMMENT{$n$ is the bin index, $t$ is the time index, $I$ is in units of $I_{sat}$} 
\STATE $\delta[n, t=0]=0$ \COMMENT{light initially resonant, $\delta$ in units of $\Gamma/2$}
\STATE $I_f=0$
\FOR[loop over time steps]{$t=0$ to $t_f$}
 \FOR[loop over bins, N is total bin number]{$n=1$ to $N$}
 \STATE $A=\sigma_0\rho[n] dz$
 \STATE $B=v_r dt/(\hbar c \rho[n])$ 
\STATE $I[n,t]=I[n-1,t] - A I[n-1,t]/(1+\delta[n,t-1]^2+I[n-1,t])$  \COMMENT{Eq. (\ref{eq3})}
\STATE $\delta[n,t]=\delta[n,t-1]+B\left(I[n-1,t]-I[n,t]\right)$  \COMMENT{Eq. (\ref{eq4})} 
\ENDFOR 
\STATE $I_f  \mathrel{+}= I[N,t]dt$
\ENDFOR
\STATE $OD_{corr1}=-\ln{(I_f/I_0t_f)}$
\end{algorithmic}
\end{algorithm}

\par To check the validity of our simulation, we can take it to certain limits in which the problem becomes analytically solvable. In the limit that the probe intensity is much weaker than saturation, $I_0\ll I_{sat}$, the atoms will not absorb enough light to significantly detune from resonance. We can then neglect time dependence and Eq.(\ref{eq3}) reduces to 
\begin{equation}
\frac{dI(z)}{dz}=-\rho\sigma_0 I(z),
\end{equation}
from which we recover the simple
\begin{equation}
\sigma_0 n = OD_0 = -\ln I_0/I_f. \label{eq6}
\end{equation}
In the limit that the probe intensity is much stronger than saturation, $I_0\gg I_sat$, even far detuned atoms will absorb light at their maximum, allowing us to again neglect the time dependence and reduce \ref{eq3} to 
\begin{equation}
\frac{dI(z)}{dz}=-\rho\sigma_0 I_{sat}, 
\end{equation}
which integrates to 
\begin{equation}
\sigma_0 n = \frac{I_0 - I_f}{I_{sat}}. \label{eq8}
\end{equation}
We recognize the right hand sides of Eq. (\ref{eq6}) and Eq. (\ref{eq8}) as the two terms in the expression for $OD_1$, Eq. (\ref{eq2}). Thus, in both limits $OD_{corr1}$ should coincide with the value of $OD_0$ obtained by fixing an input atom number $n$ in Eq. (\ref{eq2}). Indeed, they coincide as seen in Fig. \ref{fig:IsatLimits}.
\begin{figure}
	\includegraphics{figure3.pdf}
\caption{Optical density as a function of probe intensity as predicted by the simulation (blue dots) and by Eq. (\ref{eq6}) (green lines), for three different imaging times. The predictions agree in both the high and low intensity limits, and differ for probe intensities comparable to the saturation intensity. The difference is enhanced with increased imaging time.}  
\label{fig:IsatLimits}
\end{figure}
\par We can use the results of this simulation to check if the stationary atom assumption is self consistent, ie if the distance traveled by the atoms in one bin during the imaging time is less than the bin size. However, as can be seen from Fig. \ref{fig:atomTravel}, it turns out that not only do the atoms travel more than the bin size, but they also travel more than the size of the whole cloud, and the back of the cloud even overtakes the front for long enough imaging times. Thus, the atomic distribution as a function of position changes dramatically during the imaging time, and our stationary assumption is completely invalid. 
\begin{figure}
	\includegraphics{figure4.pdf}
\caption{Position of atoms as a function of imaging time for atoms in the first, middle, and last bins of the simulation. The probe intensity here is $1.2 I_{sat}$, and the optical depth is 1.6.}  
\label{fig:atomTravel}
\end{figure}

@@

To account for the changing atomic distribution during the imaging pulse, we shift our framework slightly from spatial bins of equal size with varying atoms number to superatoms. These superatoms were spatially distributed to match a typical experimental distribution, each describing the aggregate behavior of $N_{sa}$ atoms. The amended algorithm is shown in alg.[\ref{algorithm2}]. 

@@

\par First, we check the velocity behavior in this model against known limits. One such limit is that of a single supeatom. In this case, there is no attenuation, and the intensity seen by the superatom is constant at $I_0$. The only thing that evolves in time is the detuning of the single superatom. We can differentiate both sides of Eq. (\ref{eq4}) to obtain

@@

\frac{d\delta\left(t\right)}{dt}=\frac{v_r}{\hbar c \rho}\frac{dI}{dz},
\label{eq9}
and Eq. (\ref{eq3}) simplifies to
\frac{dI}{dt}=\frac{v_r \sigma_0}{\hbar c}\frac{I_0}{1+(2\delta(t)/\Gamma)^2 +I_0/I_{sat}}.
\label{oneAtom}
Combining Eq. (\ref{eq9}) and Eq. (\ref{oneAtom}), and converting to a dimensionless form, we obtain
\frac{d\Delta(t)}{dt}= k v_r \frac{\tilde{I}}{1+\Delta^2+\tilde{I}},

@@

where $\Delta = 2\delta/\Gamma$, $\tilde{I} = I_0/I_{sat}$ and $k$ is the photonic wavevector. Equation (\ref{eq10}) can be solved numerically, and is in good agreement with our simulation, as seen in Fig. \ref{fig:oneAtomVel}.

@@

\par We use this model to study the time evolution of the cloud shape during imaging. This can be visualized as the phase space evolution of superatoms, as seen in Fig. \ref{fig:phaseSpace}. We see that the cloud shape is actually strongly distorted during the imaging time. 

@@

\par We compare the optical depths predicted by each of the two models, $OD_{corr1}$ and $OD_{corr2}$. We find that, depite the significant changes in atomic distribution during the imaging time, the predicted optical densities are so slightly changed by including this effect that it is practically undetectable, as seen in Fig. \ref{fig:atomTravel}. In fact, the difference $\left|OD_{corr1}-OD_{corr2}\right|/OD_{corr1} \le 0.005$. Thus, for the purposes of deducing the atom density from experimental optical depths, simply using a stationary model is sufficient. Furthermore, the actual atomic distribution $\rho(z)$ is largely irrelevant, and the only observable is the 2-D atomic density $n=\int\rho(z)\mathrm{d}z$. 

@@

\par We first corrected each image for recoil induced detuning using the simulation described above. We simulated a look-up table of predicted optical depths as a function of atomic column density and probe intensity. From our two absorption images we knew the probe intensity and optical depth for each pixel. We then used the simulated look-up table in an inverted fashion to deduce the atomic column density. An example of the effect of this correction procedure is shown in Fig. \ref{fig:SampleCorrection}.  To improve the signal and compensate for our shot to shot number fluctuations, we took 15 equivalent images for each data point and, after correcting, added them together.

@@

Our final data is presented in Fig. \ref{fig:fittedFractions}. The red line is a best fit of the model given in Eq. (\ref{eq:fit}). The fit parameters we extracted are $\tilde{a}_{bg}=1.48e-3$, $\Delta = 1.0(1)$ mT, $B_0 = 20.247(2)$ mT, and $C=9.00e-5$. The error bars on the fitted data were obtained solely from photon shot noise of both absorption images propagated through our analysis.

>>>>

\begin{document}
\newcommand{\swave}[0]{$\it{s}$-wave }
\newcommand{\pwave}[0]{$\it{p}$-wave }
\newcommand{\K}{ $^{40}\rm{K}$}
\newcommand{\Rb}{ $^{87}\rm{Rb}$}
\newcommand{\us}{$\mu \rm{s}$} 
\newcommand{\mT}{$m\rm{T}$}

@@

\begin{abstract}
We directly measured the \swave{} scattering cross-section of \K{} atoms accross a Feshbach resonance. We collided a pair of degenerate clouds and imaged the scattered atoms. Owing to their low density, few atoms scattered, even near the resonance. To optimize signal to noise, we develop techniques to interpret absorption imaging in the regime where the optical intensity changes dramatically as light traverses the cloud, and recoil induced 
detuning corrections are significant. We applied these techniques to our \swave{} scattering data and extracted the resonant magnetic
field value. These imaging techniques are generally applicable, and can be used to observe effective \pwave{} scattering in the presence of spin-orbit coupling in a spin polarized  Fermi gas.

@@

\section{Introduction}
Feshbach resonances are a stand-by technique for tuning the interaction strength in ultracold atomic gases. Even weak interactions play a crucial role in the physics of atomic Bose-Einstein Condensates (BECs), giving rise to, for example, their characteristic Thomas-Fermi density profiles. The effects of interactions in Fermi gases, however, are harder to observe. The density of Fermi clouds is typically reduced by a factor of $~10^3$ from that of BECs, making it necessary to enhance the strength of interactions in order to observe them. Feshbach resonances alter the inter-atomic scattering length via the magnetic field. Not only do they enable detection of interactions, but they make it possible to tune interactions from attractive to repulsive, allowing for the phase transition from the BCS to BEC regime at sufficiently low temperatures \cite{RegalThesis}. 
\par A Feshbach resonance occurs when a bound molecular state of few atoms energetically approaches the free atomic state \cite{Chin10}. The relative energy of the free atomic states and the molecular state is defined a bias magnetic field. The Feshbach resonance can thus be approached by changing the bias field. The effect of the resonance on the scattering length between two free atoms in the appropriate hyperfine states is
a(B)=a_{bg}\left(1-\frac{\Delta}{B-B_0}\right),
\label{feshbachEq}
where $a$ is the scattering length, $a_{bg}$ is the scattering length away from the resonance, $\Delta$ is the width of the resonance, and $B_0$ is the field value at which the resonance occurs. Note that the scattering length tends to infinity from either side of the resonance.
\par To use Feshbach resonances as a tool it  is necessary to have a good measurement of the parameters in the above equation.  The exact value of the resonant field value $B_0$ is impossible to calculate analytically and must be estimated via numerical models \cite{Lysebo09, Gao11} or determined experimentally \cite{Inouye98, Cornish00}. Some experimental techniques that have been used to characterize Feshbach resonances include: the observation of loss due to three-body inelastic scattering, re-thermalization timescales, or anisotropic expansion, all of which infer the elastic scattering cross section from collective behavior of the cloud \cite{Regal03,OHara02,Monroe93}. 
\par Here we present a new technique for measuring the location and width of a Feshbach resonance. We collide a pair of ultra-cold Fermi gases and directly image the resulting \swave scattering halo as a function magnetic field strength. This allows us to observe the enhancement in scattering without relying on proxy effects. We measured the fraction of atoms scattered during the collision, and from this fraction deduced the resonant magnetic field  and width of the resonance.
%\par The techniques developed in this experiment for observing Fermion scattering can be extended to engineering higher order partial wave interactions, as has been done for bosons \cite{Williams2012}.
\par Our Fermi gases are so dilute that even with the resonant enhancement of the scattering cross section, only a small fraction of the atoms scatter. This makes direct detection of \swave scattering halos difficult due to detection uncertainty, which disproportionately affects regions of low atomic density. To optimize the signal to noise for low atom numbers, we utilized a non standard imaging regime. In this regime, simulation was necessary for an accurate interpretation of the images. We performed these simulations and used the results to extract the atom number and the scattered fraction from our images.
\par This paper is in two parts. In the first, we study absorption imaging in the presence of significant time-dependent Doppler shift and show how we use our results to interpret data. In the second, we describe our\swave scattering experiment and extract a measure of the location of the Feshbach resonance in \K{}.

@@

Every experiment requires a detection scheme. In ultracold atomic experiments, most measurements are based on interacting laser light with the atomic cloud. In our experments we used absorption imaging, the most commonly used technique. Absorption imaging relies on optical transitions between the ground and some selected excited atomic states. Such atomic transitions have an energy difference $E_0$ with an associated frequency $\omega_0 = E_0/\hbar$, and a natural transition linewidth $\Gamma$. When interacting with a laser light field an atom will scatter photons: it will absorb a photon from the light field and move up to its excited state, and then re-emit the photon in a random direction and decay back to its ground state. The  rate at which this scattering occurs is

@@

where $I/I_{sat}$ is the laser intensity as a fraction of the saturation intensity, and $\delta$ is the detuning, the difference between the resonant transition frequency $\omega_0$ and the frequency of the laser light $\omega_L$.  
\par An absorption image is obtained by shining an on or near resonant probe beam ($\delta\approx0$) onto the atomic cloud. Some of the light is scattered by the atoms, and the part of the beam that made it through the cloud, $I_f(x,y)$ is imaged onto a camera, as seen in Fig. \ref{fig:absorptionIntor}a (top). After the atoms leave the imaging volume, the probe light is reapplied to calibrate the intensity $I_0(x,y)$ of light unaffected by the atoms (bottom). 

@@

\caption{Absorption imaging. a. First, a near resonant probe light is shined on the atoms, and the shadow of that light is imaged on the camera. Then, the atoms are allowed to escape their trap and an image of just the probe light is taken. b.  As the probe beam, travels through the atomic cloud, some of its power is absorbed, and the intensity seen by atoms further along the imaging direction $z$ is lowered.  c. An atomic cloud illuminated by a probe light field absorbs photons from the probe and re-emits them in all directions. This process results in a net velocity of the cloud in the direction of the probe light as well as diffusive spreading in the transverse directions.  }  

@@

\par These intensities are related to the number of atoms the light encountered. Consider the light as it travels along the imaging axis $e_z$ through a 3-D atomic density profile $\rho(x,y,z)$. For the purposes of this discussion, let us focus on a single pixel of the camera, and thus a single value of $I_0$ and $I_f$ and a single column of atoms, $\rho(z)$. We will not be able to infer the entire atomic distribution, but for each pixel we can obtains the integrated density, $n = \int \rho\left(z\right) \,\mathrm{d}z$. As the light travels through a column of atoms, each atom will scatter light according to Eq. (\ref{eq:scatrate}). Therefore, the atoms further along the imaging axis $z$ will experience a reduced optical intensity, as seen in Fig. \ref{fig:absorptionIntor}b. On resonance ($\delta=0$), the intensity change from scattering as a function of $z$ is

@@

where $\sigma_0$ is the resonant scattering cross section. 
 This equation can be easily integrated to obtain  \cite{Reinaudi07}

@@

In the limit where the probe intensity is much smaller than the saturation intensity, $I_0\ll I_{sat}$, this reduces to $-\ln \left(I_f/I_0\right)$: optical depth, or $OD$. Since this is the simplest possible model that relates the observed intensities to the column density $n$, we call this quantity $OD_0$. When the probe intensity is comparable to the saturation intensity, the second term in Eq. (\ref{eq2}) becomes significant, and we define the optical density corrected for high probe intensity, or

@@

\par When an atom absorbs a photon from the laser light field and is excited to a higher energy level, by conservation of momentum it must also acquire a velocity in the direction of the light field. This recoil velocity is given by $v_r=\hbar k/m$, where $k=2\pi/\lambda$ is the wavenumber of the light field with wavelength $\lambda$ and $m$ is the atomic mass. When the atom returns to its ground state, the photon is re-emitted with some momentum $\vec{p}_e$. Over many photons, this momentum distribution averages to zero, $\int\vec{p}_e\mathrm{d}^3 x=0$.  The variance, however, is not zero, allowing the atoms to acquire some momentum transverse to the laser field. We will ignore this correction, but the effect of this on the atomic cloud is picture in Fig. \ref{fig:absorptionIntor}c. 
\par Once the atoms absorb enough photons that they acquire a substantial velocity in the direction of the light field, this velocity Doppler shift them from resonance with the light field. After scattering $N$ photons an atom acquires an average velocity $N v_r$ and an additional detuning $\delta=k N v_r$. Therefore, even if the probe beam is initially on resonance with the atomic transition, we cannot neglect the detuning term in the scattering rate as time goes on. Furthermore, this detuning varies both with imaging time $t$ and with distance along the propagation direction $z$ (Fig. \ref{fig:detunedBlobs}). Thus, the intensity lost to the atoms also acquires a time dependence: 

@@

\frac{dI(t,z)}{dz}=\sigma_0 \rho \frac{I(t,z)}{1+[2\delta(t,z)/\Gamma]^2 +I(t,z)/I_{sat}}, \label{eq3}

@@

where the detuning $\delta$ is

@@

\delta(t,z)=\frac{v_r}{\hbar c \rho}\int_0^t \frac{dI(z,\tau)}{dz}\,\mathrm{d}\tau; \label{eq4} 

@@

the relationship between the atomic density and the observed intensities is no longer straightforward.

@@

	\includegraphics*{Figure1.pdf}
\caption{Distribution of generalized detuning $D=2\delta/\Gamma$ across an atomic cloud of \K{} for three different imaging times, as obtained by numerical simulation.}  

@@

\par By considering this equation perturbatively in time we obtain corrections to second order in imaging time \cite{LJLthesis}
\sigma_0n\approx OD^{(0)}+OD^{(1)}t+OD^{(2)}t^2 = OD_2
OD_2= OD_1+\frac{(kv_rt)^2}{3}\left[\frac{I_{sat}}{I_f+I_{sat}}-\frac{I_{sat}}{I_0+I_{sat}}+\mathrm{ln}\left(\frac{I_f+I_{sat}}{I_0+I_{sat}}\right)\right].
 However, as shown in  Fig. \ref{fig:ODcorrections}, the perturbative treatment breaks down shortly after the zeroth order approximation of Eq. (\ref{eq2}).To adequately correct for the recoil induced detuning of the atoms, we numerically simulated the imaging process for $I_f$ as a function of imaging time, atomic density, and probe intensity. 

@@

\par In the following, we describe two versions of this simulation. First, we take a simplistic approach where the spatial distribution of atoms does not change appreciably during the imaging time and can be treated statically. We test this approach in known limits and then check the validity of the static assumption.For realistic input parameters, this assumption is invalid. We developed a slightly more sophisticated quasi-classical approach and allowed the atoms to move during the imaging time, allowing us to simulate the phase space evolution of atoms subjected to probe light. While the atomic trajectories are wildly different than in the static atom approximation, the predicted ODs only vary on the 0.5$\%$ level between the two models.

\subsection{Stationary atom model}
To solve Eqs. (\ref{eq3})-(\ref{eq4}), we start with a 1-d distribution distribution of atomic densities $\rho(z)$, assumed to be Gaussian in shape. We divided the cloud into spatial bins (the bin size was chosen such that decreasing the bin size further changed the result at the $10^{-5}$ level).  In this approximation, we keep the number of atoms in each bin constant.  The algorithm used is shown in Alg. [\ref{algorithm1}]. We call the optical depth obtained from this algorithm $OD_{corr1}$.

\begin{algorithm}
\caption{Stationary atom model}
\label{algorithm1}
\begin{algorithmic}
\STATE $I[n=0,t]=I_0$ \COMMENT{$n$ is the bin index, $t$ is the time index, $I$ is in units of $I_{sat}$} 
\STATE $\delta[n, t=0]=0$ \COMMENT{light initially resonant, $\delta$ in units of $\Gamma/2$}
\STATE $I_f=0$ \COMMENT{no light has entered the cloud}
\FOR[loop over time steps]{$t=0$ to $t_f$}
 \FOR[loop over bins, N is total bin number]{$n=1$ to $N$}
 \STATE $A=\sigma_0\rho[n] dz$ \COMMENT{$dz$ is the size of spatial step}
 \STATE $B=v_r dt/(\hbar c \rho[n])$  \COMMENT{$dt$ is the size of the time step}
\STATE $I[n,t]=I[n-1,t] - A I[n-1,t]/(1+\delta[n,t-1]^2+I[n-1,t])$  \COMMENT{Eq. (\ref{eq3})}
\STATE $\delta[n,t]=\delta[n,t-1]+B\left(I[n-1,t]-I[n,t]\right)$  \COMMENT{Eq. (\ref{eq4})} 
\ENDFOR 
\STATE $I_f =I_f+ I[N,t]dt$
\ENDFOR
\STATE $OD_{corr1}=-\ln{(I_f/I_0t_f)}$
\end{algorithmic}
\end{algorithm}

\par We checked the validity of our simulation in the limits in were the problem is analytically solvable. When the probe intensity is much weaker than the saturation intensity, $I_0\ll I_{sat}$, the atoms' velocity is hardly changed, and Eq.(\ref{eq3}) reduces to 
\begin{equation}
\frac{dI(z)}{dz}=-\rho\sigma_0 I(z),
\end{equation}
from which we recover 
\begin{equation}
\sigma_0 n = OD_0 = -\ln I_0/I_f. \label{eq6}
\end{equation}
In the limit that the probe intensity is much stronger than the saturation intensity, $I_0/I_{sat}\gg \delta/2\Gamma$, even far detuned atoms will absorb light at their maximum, allowing us to again neglect the time dependence of the detuning. Equation (\ref{eq3}) becomes 
\begin{equation}
\frac{dI(z)}{dz}=-\rho\sigma_0 I_{sat}, 
\end{equation}
which integrates to 
\begin{equation}
\sigma_0 n = \frac{I_0 - I_f}{I_{sat}}. \label{eq8}
\end{equation}
We recognize the right hand sides of Eq. (\ref{eq6}) and Eq. (\ref{eq8}) as the two terms in the expression for $OD_1$ in Eq. (\ref{eq2}). Thus, as shown in  Fig. \ref{fig:IsatLimits}, in both limits $OD_{corr1}$  coincides with $OD_0$ in both the small and large probe intensity limits. 
\begin{figure}
	\includegraphics{figure3.pdf}
\caption{Optical density as a function of probe intensity as predicted by the simulation (blue dots) and by Eq. (\ref{eq6}) (green lines), for three different imaging times. The predictions agree in both the high and low intensity limits, and differ for probe intensities comparable to the saturation intensity. The difference is enhanced with increased imaging time.}  
\label{fig:IsatLimits}
\end{figure}
\par We used the results of this simulation to check if the stationary atom assumption is valid, i.e. if the distance traveled by the atoms during the imaging time is less than the bin size. As can be seen from Fig. \ref{fig:atomTravel}, not only do the atoms travel more than the bin size, but they travel far beyond the initial extent of the cloud. Moreover, owing to the high scatter rate, the back of the cloud overtakes the front for long imaging times. Thus, the atomic distribution as a function of position changes dramatically during the imaging pulse, and the stationary assumption is invalid. 
\begin{figure}
	\includegraphics{figure4.pdf}
\caption{Position of atoms as a function of imaging time for atoms in the first, middle, and last bins of the simulation. The probe intensity here is $1.2 I_{sat}$, and the optical depth is 1.6.}  
\label{fig:atomTravel}
\end{figure}

@@

To account for the changing atomic distribution during the imaging pulse, we simulate the classical kinetics of atoms subject to the recoil driven optical forces. To simulate large ensembles in a reasonable time, we use "superatoms",  each describing the aggregate behavior of $N_{sa}$ atoms. The amended algorithm is shown in alg.[\ref{algorithm2}]. 

@@

\par First, we check the velocity predicted in this model against known limits. One such limit is that of a single superatom. In this case, there is no attenuation, and the intensity seen by the superatom is constant at $I_0$. Only the detuning of the single superatom evolves in time, and Eqs. (\ref{eq3}) and (\ref{eq4}) give

@@

\frac{dD(t)}{dt}= k_R v_r \frac{\tilde{I}}{1+D^2+\tilde{I}},

@@

where $D = 2\delta/\Gamma$, and $\tilde{I} = I_0/I_{sat}$. Equation (\ref{eq10}) can be solved numerically, and is in good agreement with our simulation, as seen in Fig. \ref{fig:oneAtomVel}.

@@

\par We use this model to study the time evolution of the cloud shape during imaging and visualized the phase space evolution of superatoms, shown in Fig. \ref{fig:phaseSpace}. The cloud shape is strongly distorted during the imaging time. 

@@

\par We compare the optical depths predicted by each of the two models, $OD_{corr1}$ and $OD_{corr2}$, and, as seen Fig. \ref{fig:atomTravel}, the predicted optical depths are hardly changed by including the full time evolution:  $\left|OD_{corr1}-OD_{corr2}\right|/OD_{corr1} \le 0.005$. Thus, for the purposes of deducing the atom density from experimental optical depths,the stationary atom model is sufficient. Furthermore, we simulated a range of initial density profiles $\rho(z)$, and found their impact to be negligible - the only observable is the integrated atomic density $n=\int\rho(z)\mathrm{d}z$. 

@@

\par We first corrected each image for recoil induced detuning as described in the previous section. An example of this correction procedure is shown in Fig. \ref{fig:SampleCorrection}.  To improve the signal and compensate for our shot to shot number fluctuations, we took 15 nominally identical images for each data point and, after correcting, averaged them.

@@

Our final data is presented in Fig. \ref{fig:fittedFractions}. The red curve depicts a best fit of the model given in Eq. (\ref{eq:fit}). The fit parameters we extracted were $\Delta = 1.0(1)$  \mT{}, $B_0 = 20.247(2)$  \mT{}, and $C=9.00e-5$. The error bars on the fitted data were obtained solely from photon shot noise of both absorption images propagated through our analysis.
